### 2021-06-15 submission/manuscript.Rmd
Address edits from Pat

---
title: ''
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---

@@

## Abstract
Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-challenge (dpc). PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remain colonized for up to 30 dpc. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpc. To examine how PEG treatment impacts clearance, we administered a 1-day PEG treatment to clindamycin-treated, *C. difficile*-challenged mice. Administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpc. When we trained a random forest model with community data from 5 dpc, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90). Examining the dynamics of these bacterial during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C. difficile* colonization in PEG-treated mice.

## Importance
Diarrheal samples induced by medications such as laxatives are typically rejected for *Clostridiodes difficile* testing. However, there are some microbiota similarities between diarrheal and *C. difficile* infection (CDI) communities such as lower diversity compared to samples from healthy patients, which lead us to hypothesize that diarrhea may be an indicator of *C. difficile* risk. We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C. difficile* by administering a laxative to mice either before or after *C. difficile* challenge. Our findings suggest the osmotic laxative disrupts colonization resistance to *C. difficile*, as well as clearance in mice already colonized with *C. difficile*. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative samples from patients are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and *C. difficile* negative patients with diarrhea compared to samples that were *C. difficile* negative with non-diarrheal stool consistency [@Schubert2014]. These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.

Osmotic laxatives can lead to diarrhea depending on the dose administered and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization resistance. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives [@vanNood2013; @Razik2017; @Postigo2012] based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

Our group has used C57BL/6 mice to characterize how antibiotics disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance. Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C. difficile* colonization and severe CDI than mice treated with clindamycin, and if PEG treatment after challenge can promote *C. difficile* clearance. Addressing these questions will better inform how we think about laxatives and diarrhea in the context of CDIs.

## Results
**5-day laxative treatment led to prolonged *C. difficile* colonization in mice.** Building off of previous work that showed treating mice with the osmotic laxative, PEG 3350, rendered mice susceptible to *C. difficile* colonization [@Ferreyra2014; @VanInsberghe2020], we decided to test how long *C. difficile* colonization is sustained and how long PEG-treated mice remain susceptible to *C. difficile*.  We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile* colonization, with mice typically clearing *C. difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020]. All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days. The first group received no additional treatment. The second group was also treated with clindamycin. A third group was allowed to recover for 10 days prior to challenge (Fig. 1A). The PEG treatment resulted in weight loss for the 3 groups of mice, with the greatest change in weight observed on the fifth day of the PEG treatment. The mice recovered most of the lost weight by five days after treatment (Fig. 1B). After either the PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. In contrast to the mice that only received clindamycin, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-challenge (Fig. 1C). The clindamycin-treated mice cleared *C. difficile* within ten days post-challenge (Fig. 1C). It was noteworthy that PEG-treated mice were still susceptible to *C. difficile* colonization after a 10-day recovery period, although *C. difficile* was not detectable in most of the group in the initial five days post-challenge (Fig. 1C, S1A). One mouse was found dead on the 6th day post-challenge, presumably due to *C. difficile* as the bacterium became detectable in stool samples from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward, the density of *C. difficile* stabilized in the 10-day recovery group and remained high through 20-30 days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible through ten days post-treatment.

@@

Since none of the communities completely recovered in the follow-up period after treatments, we next profiled community diversity and composition. We examined the alpha diversity dynamics by calculating the communities' Shannon diversity. Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-challenge (Fig. 2C; Data Set S1, sheet 2). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment. We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, sheet 3). The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased. The increase in *Bacteroides* relative abundance was unique to PEG treated mice, as the *Bacteroides* relative abundance actually decreased in clindamycin treated mice (Fig. 2D). Finally, we identified the genera whose relative abundance differed across treatment groups over multiple time points. Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig. 2E, Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

@@

Because there were differences in the mucosal microbiota including detectable *C. difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C. difficile* challenge by evaluating cecum and colon histopathology [@Reeves2011]. However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although there was a slight difference in the histopathology score of the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5 to 2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* 630 challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

```
**Five-day post-challenge community data can predict which mice will have prolonged *C. difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG treatments, we examined the bacteria that influenced prolonged *C. difficile* colonization. We trained 3 machine learning models (random forest, logistic regression, and support vector machine) with bacterial community data from 5 days post-challenge to predict whether the mice were still colonized with *C. difficile* 10 days post-challenge. We chose 5 days post-challenge because that was the earliest time point where we saw a treatment effect in the post-challenge 1-day PEG experiments. The random forest model had the highest performance (median AUROC = `r max_auc`, Data Set S1, sheet 19), suggesting the 5-day post challenge microbiota is an excellent predictor of prolonged *C. difficile* colonization. Next, we performed permutation importance to examine the bacteria that were the top contributors to the random forest model predicting prolonged *C. difficile *colonization. We selected 10 genera that contributed the most to our model's performance (Fig. 7A) and examined their relative abundance at 5 days post-challenge, the time point used to predict *C. difficile* colonization status on day 10 (Fig. 7B). Next, we focused on the 5 genera that had a greater than 1% relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time. We found *Enterobacteriaceae* and *Bacteroides* tended to have a higher relative abundance, the relative abundance of *Akkermansia* was initially decreased and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C. difficile* (Fig. 7C). Together these results suggest a combination of low and high abundance bacterial genera influence the prolonged colonization observed in 5-day PEG and post-challenge 1-day PEG treated mice.

## Discussion
While the disruptive effect of antibiotics on *C. difficile* colonization resistance is well established, the extent to which other drugs such as laxatives disrupt colonization resistance was unclear. By following osmotic laxative treated mice over time, we found that a 5-day PEG treatment before challenge resulted in prolonged *C. difficile* colonization, while a 1-day PEG treatment resulted in transient colonization. The differences in *C. difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in how much the treatments disrupted the microbiota. Additionally, the intestinal communities of 5-day PEG treated mice did not regain colonization resistance after a 10-day recovery period. In contrast to the other 5-day PEG treatment groups, *C. difficile* was not immediately detectable in the stools of most of the mice in the 10-day recovery group. We also examined the impact of PEG treatment after *C. difficile* challenge. In opposition to the hypothesis suggested by the literature, we found that PEG treatment prolonged colonization relative to mice that only recieved clindamycin treatment. We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C. difficile* colonization (Fig. 8). Overall, our results demonstrated that osmotic laxative treatment alone rendered mice susceptible to *C. difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after challenge.

@@

Considering laxatives are also used to prepare patients when administering fecal microbiota transplants via colonoscopy to treat recurrent CDIs, it will be important to determine if osmotic laxatives impact *C. difficile* clearance in the human intestinal tract. It is still unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs [@Kampouri2021]. Nevertheless, results from the FMT National Registry where 85% of FMTs were delivered by colonoscopy demonstrate FMTs are highly effective treatment for recurrent CDIs with 90% achieving resolution in the 1 month follow-up window [@Kelly2021]. A surprising number of studies continue to hypothesize that PEG or bowel preparation can clear *C. difficile* spores and toxins despite the paucity of supporting evidence [@Liacouras1996; @vanNood2013; @Razik2017; @Postigo2012]. There was even a clinical trial (NCT01630096) designed to examine whether administering PEG 3350 (NuLYTELY) prior to antibiotic treatment reduced disease severity that started recruitment in 2012  [@Dieterle2018], but no results have been posted to date. Here we sought to evaluate the impact of treating *C. difficile* colonized mice with PEG (with or without FMT) and found clearance was delayed. Further studies are needed to understand the impact of osmotic laxatives on *C. difficile* colonization resistance and clearance in human patients receiving FMTs.

@@

***C. difficile* challenge model.** Mice were challenged with 25 microliters off *C. difficile* 630 spores at 10^5^ concentration, except for 1 experiment where the concentration was 10^3^ (Fig. 5A). All mock challenged mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper pipette was used to administer the same challenge dose to mice via oral gavage. Mice were weighed daily throughout the experiment and stool was collected for quantifying *C. difficile* CFU and 16S rRNA gene sequencing. There were two groups of mice that received either a PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal microbiota transplant was prepared with stool samples collected from the mice in the experiment prior to the start of any treatments. The stool samples were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS and glycerol was added to make a 15% glycerol solution. The solution was then aliquoted into tubes and stored at -80&deg;C until the day of the gavage. An aliquot of both the FMT and PBS solutions were also set aside in the -80&deg;C for 16S rRNA gene sequencing. The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The supernatant was then transferred to a separate tube to prevent the gavage needle from clogging with debris during gavage. The PBS solution that was administered to the other group was also 15% glycerol. Each mouse was administered 100 microliters of either the FMT or PBS solution via gavage. When we refer to mice that cleared *C. difficile*, we mean that no *C. difficile* was detected in the first serial dilution (limit of detection: 100 CFU). In some experiments, we collected tissues for 16SrRNA gene sequencing, histopathology, or both. For 16S rRNA gene sequencing, we collected small snips of cecum, proximal colon, and distal colon tissues in microcentrifuge tubes, snap froze in liquid nitrogen, and stored at -80&deg;C. For histopathology, cecum and colon tissues were placed into separate casettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.

@@

**16S rRNA gene sequencing.** Stool samples that were stored in the -80&deg;C were placed into 96-well plates for DNA extractions and library preparation. DNA was extracted using the DNeasy Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf). For library preparation, each plate had a mock community control (ZymoBIOMICS microbial community DNA standards) and a negative control (water). The V4 region of the 16S rRNA gene was amplified with the AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom barcoded primers, as previously described [@Kozich2013]. The PCR amplicons were normalized (SequalPrep normalizatin plate kit from Thermo Fisher Scientific), pooled and quantified (KAPA library quantification kit from KAPA Biosystems), and sequenced with the MiSeq system (Illumina).

>>>>

---
title: ''
csl: msphere.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
  html_document:
    df_print: paged
geometry: margin=1.0in
bibliography: references.bib
---

@@

## Abstract
Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350 also alter the microbiota. But whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-challenge. PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remained colonized for up to 30 days post-challenge. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 days post-challenge. To examine how PEG treatment impacts clearance, we administered a 1-day PEG treatment to clindamycin-treated, *C. difficile*-challenged mice. Administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 days post-challenge. When we trained a random forest model with community data from 5 days post-challenge, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90). Examining the dynamics of these bacterial populations during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C. difficile* colonization in PEG-treated mice. Thus, the osmotic laxative, PEG, rendered mice susceptible to *C. difficile* colonization and hindered clearance.

## Importance
Diarrheal samples from patients taking laxatives are typically rejected for *Clostridiodes difficile* testing. However, there are similarities between the bacterial communities from people with diarrhea or *C. difficile* infections (CDI) including lower diversity compared to communities from healthy patients, which led us to hypothesize that diarrhea may be an indicator of *C. difficile* risk. We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C. difficile* by administering a laxative to mice either before or after *C. difficile* challenge. Our findings suggest that osmotic laxatives disrupt colonization resistance to *C. difficile*, and prevent clearance among mice already colonized with *C. difficile*. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative samples from patients are separated into two groups based on stool consistency, there are similar microbiota features between samples from CDI patients and *C. difficile* negative patients with diarrhea compared to samples that were *C. difficile* negative with non-diarrheal stool consistency [@Schubert2014]. The similar community features between CDI patients and patients with diarrhea included low alpha diversity and only 6 bacterial taxa had higher relative abundances in communities from CDI patients. These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.

depending on the dose administered, osmotic laxatives can lead to diarrhea and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization resistance. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives prior to the FMT [@vanNood2013; @Razik2017; @Postigo2012] based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

Our group has used C57BL/6 mice to characterize how antibiotics disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although two groups have now shown that PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance. Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C. difficile* colonization and severe CDI than mice treated with clindamycin, and whether PEG treatment after challenge can promote *C. difficile* clearance. Addressing these questions will better inform how we think about laxatives and diarrhea in the context of CDIs.

## Results
**5-day laxative treatment led to prolonged *C. difficile* colonization in mice.** Building off of previous work that showed treating mice with the osmotic laxative, PEG 3350, rendered mice susceptible to *C. difficile* colonization [@Ferreyra2014; @VanInsberghe2020], we decided to test how long *C. difficile* colonization is sustained and how long PEG-treated mice remain susceptible to *C. difficile*.  We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders mice susceptible to *C. difficile* colonization, with mice typically clearing *C. difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020]. All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days. The first group received no additional treatment. The second group was also treated with clindamycin. A third group was allowed to recover for 10 days prior to challenge (Fig. 1A). The PEG treatment resulted in weight loss for the 3 groups of mice, with the greatest change in weight observed on the fifth day of the PEG treatment. The mice recovered most of the lost weight by five days after treatment (Fig. 1B). After either the PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. In contrast to the mice that only received clindamycin, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-challenge (Fig. 1C). The clindamycin-treated mice cleared *C. difficile* within ten days post-challenge (Fig. 1C). It was noteworthy that PEG-treated mice were still susceptible to *C. difficile* colonization after a 10-day recovery period, although *C. difficile* was not detectable in most of the group in the initial five days post-challenge (Fig. 1C, S1A). One mouse was found dead on the 6th day post-challenge, presumably due to *C. difficile* as the bacterium became detectable in stool samples from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward, the density of *C. difficile* stabilized in the 10-day recovery group and remained high through 20-30 days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible through ten days post-treatment.

@@

Because time and treatment group influenced most of the variation between communities, we next profiled community diversity and composition. We examined the alpha diversity dynamics by calculating the communities' Shannon diversity. Although both clindamycin and PEG treatments decreased diversity, the Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin alone through thirty days post-challenge (Fig. 2C; Data Set S1, sheet 2). We next identified the bacterial genera whose relative abundances shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment. We found `r PEG_genera` genera whose relative abundances were altered by PEG treatment (Data Set S1, sheet 3). The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased. The increase in *Bacteroides* relative abundance was unique to PEG treated mice, as the *Bacteroides* relative abundance actually decreased in clindamycin treated mice (Fig. 2D). Finally, we identified the genera whose relative abundance differed across treatment groups over multiple time points. Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig. 2E, Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

@@

Because there were differences in the mucosal microbiota including detectable *C. difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C. difficile* challenge by evaluating cecum and colon histopathology [@Reeves2011]. However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although there was a slight difference in the histopathology score of the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5 to 2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

@@

```
**Five-day post-challenge community data can predict which mice will have prolonged *C. difficile* colonization.** After identifying bacteria associated with the 5-day, 1-day and post-challenge 1-day PEG treatments, we examined the bacteria that influenced prolonged *C. difficile* colonization. We trained 3 machine learning models (random forest, logistic regression, and support vector machine) with bacterial community data from 5 days post-challenge to predict whether the mice were still colonized with *C. difficile* 10 days post-challenge. We chose to predict the status based on communities 5 days post-challenge because that was the earliest time point where we saw a treatment effect in the post-challenge 1-day PEG experiments. The random forest model had the highest performance (median AUROC = `r max_auc`, Data Set S1, sheet 19) and indicated that the 5-day post challenge microbiota was an excellent predictor of prolonged *C. difficile* colonization. Next, we performed a permutation importance test to identify the bacteria that were the top contributors to the random forest model for predicting prolonged *C. difficile *colonization. We selected 10 genera that contributed the most to our model's performance (Fig. 7A) and examined their relative abundance at 5 days post-challenge, the time point used to predict *C. difficile* colonization status on day 10 (Fig. 7B). Next, we focused on the 5 genera that had a greater than 1% relative abundance in either the cleared or colonized mice and examined how the bacteria changed over time. We found *Enterobacteriaceae* and *Bacteroides* tended to consistently have a higher relative abundance, the relative abundance of *Akkermansia* was initially low and then increased, and *Porphyromonadaceae* and *Lachnospiraceae* had a lower relative abundance in the mice with prolonged colonization compared to the mice that cleared *C. difficile* (Fig. 7C). Together these results suggest a combination of low and high abundance bacterial genera influence the prolonged colonization observed in 5-day PEG and post-challenge 1-day PEG treated mice.

## Discussion
While the disruptive effect of antibiotics on *C. difficile* colonization resistance is well established, the extent to which other drugs such as laxatives disrupt colonization resistance was unclear. By following mice treated with an osmotic laxative over time, we found that a 5-day PEG treatment before challenge resulted in prolonged *C. difficile* colonization, while a 1-day PEG treatment resulted in transient colonization without the use of antibiotics. The differences in *C. difficile* colonization dynamics between the 5- and 1-day PEG treated mice were associated with differences in how much the treatments disrupted the microbiota. Additionally, the intestinal communities of 5-day PEG treated mice did not regain colonization resistance after a 10-day recovery period. In contrast to the other 5-day PEG treatment groups, *C. difficile* was not immediately detectable in the stools of most of the mice in the 10-day recovery group. We also examined the impact of PEG treatment after *C. difficile* challenge. In opposition to the hypothesis suggested by the literature, we found that PEG treatment prolonged colonization relative to mice that only recieved clindamycin treatment. We identified patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Akkermansia*, *Porphyromonadaceae*, and *Lachnospiraceae* that were associated with prolonged *C. difficile* colonization (Fig. 8). Overall, our results demonstrated that osmotic laxative treatment alone rendered mice susceptible to *C. difficile* colonization and the duration of colonization depended on the length of PEG treatment and whether treatment was administered before or after challenge.

@@

Considering laxatives are also used to prepare patients when administering fecal microbiota transplants via colonoscopy to treat recurrent CDIs, it will be important to determine whether osmotic laxatives impact *C. difficile* clearance in the human intestinal tract. It is still unclear what the best administration route is because there have been no studies designed to evaluate the best administration route for FMTs [@Kampouri2021]. Nevertheless, results from the FMT National Registry where 85% of FMTs were delivered by colonoscopy demonstrate FMTs are highly effective treatment for recurrent CDIs with 90% achieving resolution in the 1 month follow-up window [@Kelly2021]. A surprising number of studies continue to hypothesize that PEG or bowel preparation can clear *C. difficile* spores and toxins despite the paucity of supporting evidence [@Liacouras1996; @vanNood2013; @Razik2017; @Postigo2012]. There was even a clinical trial (NCT01630096) designed to examine whether administering PEG 3350 (NuLYTELY) prior to antibiotic treatment reduced disease severity that started recruitment in 2012  [@Dieterle2018], but no results have been posted to date. Here we sought to evaluate the impact of treating *C. difficile* colonized mice with PEG (with or without FMT) and found clearance was delayed. Further studies are needed to understand the impact of osmotic laxatives on *C. difficile* colonization resistance and clearance in human patients receiving FMTs.

@@

***C. difficile* challenge model.** Mice were challenged with 25 microliters of *C. difficile* 630 spores at 10^5^ concentration, except for 1 experiment where the concentration was 10^3^ (Fig. 5A). All mock challenged mice received 25 ul vehicle solution (Ultrapure water). A Dymax stepper pipette was used to administer the same challenge dose to mice via oral gavage. Mice were weighed daily throughout the experiment and stool was collected for quantifying *C. difficile* CFU and 16S rRNA gene sequencing. There were two groups of mice that received either a PBS or fecal microbiota transplant (FMT) gavage post-PEG treatment. The fecal microbiota transplant was prepared with stool samples collected from the mice in the experiment prior to the start of any treatments. The stool samples were transferred to an anaerobic chamber and diluted 1:10 in reduced PBS and glycerol was added to make a 15% glycerol solution. The solution was then aliquoted into tubes and stored at -80&deg;C until the day of the gavage. An aliquot of both the FMT and PBS solutions were also set aside in the -80&deg;C for 16S rRNA gene sequencing. The day of the gavage, aliquots were thawed and centrifuged at 7500 RPM for 1 minute. The supernatant was then transferred to a separate tube to prevent the gavage needle from clogging with debris during gavage. The PBS solution that was administered to the other group was also 15% glycerol. Each mouse was administered 100 microliters of either the FMT or PBS solution via gavage. When we refer to mice that cleared *C. difficile*, we mean that no *C. difficile* was detected in the first serial dilution (limit of detection: 100 CFU). In some experiments, we collected tissues for 16SrRNA gene sequencing, histopathology, or both. For 16S rRNA gene sequencing, we collected small snips of cecum, proximal colon, and distal colon tissues in microcentrifuge tubes, snap froze in liquid nitrogen, and stored at -80&deg;C. For histopathology, cecum and colon tissues were placed into separate casettes, fixed, and then submitted to McClinchey Histology Labs (Stockbridge, MI) for processing, embedding, and hematoxylin and eosin (H&E) staining.

@@

**16S rRNA gene sequencing.** Stool samples were stored at -80&deg;C and were placed into 96-well plates for DNA extractions and library preparation. DNA was extracted using the DNeasy Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system (Eppendorf). For library preparation, each plate had a mock community control (ZymoBIOMICS microbial community DNA standards) and a negative control (water). The V4 region of the 16S rRNA gene was amplified with the AccuPrime Pfx DNA polymerase (Thermo Fisher Scientific) using custom barcoded primers, as previously described [@Kozich2013]. The PCR amplicons were normalized (SequalPrep normalizatin plate kit from Thermo Fisher Scientific), pooled and quantified (KAPA library quantification kit from KAPA Biosystems), and sequenced with the MiSeq system (Illumina).

**16S rRNA gene sequence analysis.** All sequences were processed with mothur (v. 1.43) using a previously published protocol [@Schloss2009; @Kozich2013]. Paired sequencing reads were combined and  aligned with the SILVA (v. 132) reference database [@Quast2012] and taxonomy was assigned with a modified version of the Ribosomal Database Project reference sequences (v. 16) [@Cole2013]. The error rate for are sequencing data was 0.0559% based on the 17 mock communities we ran with the samples. Samples were rarefied to 1,000 sequences, 1,000 times for alpha and beta diversity analyses in order to account for uneven sequencing across samples. All but 3 of the 17 water controls had fewer than 1000 sequences. PCoAs were generated based on Bray-Curtis Index distance matrices. Permutational multivariate analysis of variance (PERMANOVA) tests were performed on mothur-generated Bray-Curtis distance matrices with the adonis function from the vegan R package [@Vegan2018].
